---
title: "*T. tana* Analyses for Thesis"
author: "Maya Juman"
date: "March 8, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#load relevant packages
library(PerformanceAnalytics)
library(ggplot2)
library(ggfortify)
library(MASS)
library(lme4)
library(car)

#custom color palette bc defaults are ugly
cbp2 <- c("black", "white", "cyan", "chartreuse", "firebrick1", "darkviolet", "gold", "forestgreen", "lightpink", "burlywood4", "darkorange", "blue", "magenta")
```


```{r}
#set up data
setwd("~/Desktop/thesis/complete data")
x <- read.csv("manus.csv")
x <- x[,-23]
y <- read.csv("cran.csv")

#scaling by CPL
yCPL <- y
yCPL <- yCPL[!(is.na(yCPL$CPL)),]

for (j in 1:169) {
  for (i in 10:27) {
  yCPL[j,i] <- (yCPL[j,i])/(yCPL$CPL[j])
} 
} 

yCPL <- na.omit(yCPL[,-c(9:11,14,17:20,23,26,27)])

x[,7:22] <- log(x[,7:22])
y[,9:27] <- log(y[,9:27])

x <- x[,colSums(is.na(x)) < 20]
x <- na.omit(x)

y <- y[,colSums(is.na(y)) < 20]
y <- na.omit(y)
```

##All Subspecies PCAs

###1. All Subspecies PCA (Manus)

```{r}
pc1 <- princomp(x[,-(1:6)], cor=TRUE)
for (i in 1:8) {
  pc1$loadings[,i] <- (pc1$loadings[,i] * pc1$sdev[i])
}
print(summary(pc1),digits=4,loadings=pc1$loadings,cutoff=0)
autoplot(pc1, data = x, colour = 'Subspecies', label = FALSE, label.size = 3) + scale_color_manual(values = cbp2) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (50.66%)") + ylab("PC2 (15.35%)")
autoplot(pc1, data = x, colour = 'Locality', label = FALSE, 
         label.size = 3, frame = TRUE) + xlab("PC1 (50.66%)") + ylab("PC2 (15.35%)")
round(pc1$sdev^2,2)
```

###2. All Subspecies PCA (Skull)

```{r}
pc2 <- princomp(y[,-(1:8)], cor=TRUE)
for (i in 1:8) {
  pc2$loadings[,i] <- (pc2$loadings[,i] * pc2$sdev[i])
}
print(summary(pc2),digits=2,loadings=pc2$loadings,cutoff=0)
autoplot(pc2, data = y, colour = 'Subspecies', label = FALSE, label.size = 3) + scale_color_manual(values = cbp2) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (65.58%)") + ylab("PC2 (13.92%)")
autoplot(pc2, data = y, colour = 'Locality', label = FALSE, 
         label.size = 3, frame = TRUE) + xlab("PC1 (65.58%)") + ylab("PC2 (13.92%)") + ggtitle("Original skull PCA with all individuals")
round(pc2$sdev^2,2)
```

```{r}
#scaling with CPL
pc2cpl <- princomp(yCPL[,-(1:8)], cor=TRUE)
for (i in 1:8) {
  pc2cpl$loadings[,i] <- (pc2cpl$loadings[,i] * pc2cpl$sdev[i])
}
print(summary(pc2cpl),digits=2,loadings=pc2cpl$loadings,cutoff=0)
autoplot(pc2cpl, data = yCPL, colour = 'Subspecies', label = FALSE, label.size = 3) + scale_color_manual(values = cbp2) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (29.21%)") + ylab("PC2 (25.71%)")
autoplot(pc2cpl, data = yCPL, colour = 'Locality', label = FALSE, 
         label.size = 3, frame = TRUE) + xlab("PC1 (29.21%)") + ylab("PC2 (25.71%)") + ggtitle("Skull PCA with all individuals, scaled with CPL")
round(pc2cpl$sdev^2,2)
```


##Mainland Borneo PCAs

###3. Mainland Borneo PCA (Manus)

```{r}
bornX <- x[which(x$Locality == 'Mainland Borneo'),]
pc3 <- princomp(bornX[,-(1:6)], cor=TRUE)
for (i in 1:8) {
  pc3$loadings[,i] <- (pc3$loadings[,i] * pc3$sdev[i])
}
print(summary(pc3),digits=2,loadings=pc3$loadings,cutoff=0)
round(pc3$sdev^2,2)

autoplot(pc3, data = bornX, colour = 'Subspecies', label = FALSE, label.size = 3) + scale_color_manual(values = cbp2) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (54.63%)") + ylab("PC2 (12.56%)")

autoplot(pc3, data = bornX, colour = 'Subspecies', label = FALSE, label.size = 3, frame = TRUE) + xlab("PC1 (54.63%)") + ylab("PC2 (12.56%)")

#table(bornX$Subspecies)
```

###4. Mainland Borneo PCA (Skull)

```{r}
bornY <- y[which(y$Locality == 'Mainland Borneo'),]
pc4 <- princomp(bornY[,-(1:8)], cor=TRUE)
for (i in 1:8) {
  pc4$loadings[,i] <- (pc4$loadings[,i] * pc4$sdev[i])
}
print(summary(pc4),digits=2,loadings=pc4$loadings,cutoff=0)
round(pc4$sdev^2,2)

autoplot(pc4, data = bornY, colour = 'Subspecies', label = FALSE, label.size = 3) + scale_color_manual(values = cbp2) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (58.52%)") + ylab("PC2 (17.50%)")

autoplot(pc4, data = bornY, colour = 'Subspecies', label = FALSE, label.size = 3, frame = TRUE) + xlab("PC1 (58.52%)") + ylab("PC2 (17.50%)")
```

##Borneo vs. Sumatra *T. t. tana* DFAs

###5. Borneo vs. Sumatra *T. t. tana* DFA (Manus)

```{r}
tanaX <- x[which(x$Subspecies == 'T. t. tana'),]

tanaX$Locality <- droplevels(tanaX$Locality)

lin <- lda(tanaX[,-c(1:6)],grouping=tanaX$Locality)
ctraw <- table(tanaX$Locality, predict(lin)$class)
ctraw #rows are actual count and columns are predicted count; ignore blank I column/row

# total percent correct
round(sum(diag(prop.table(ctraw))),4)

table(tanaX$Locality)
```

###6. Borneo vs. Sumatra *T. t. tana* DFA (Skull)

```{r}
tanaY <- y[which(y$Subspecies == 'T. t. tana'),]

tanaY$Locality <- droplevels(tanaY$Locality)

table(tanaY$Locality)

lin2 <- lda(tanaY[,-c(1:8)],grouping=tanaY$Locality)
ctraw2 <- table(tanaY$Locality, predict(lin2)$class)
ctraw2 #rows are actual count and columns are predicted count; ignore blank I column/row

# total percent correct
round(sum(diag(prop.table(ctraw2))),4)
```

##Mainland Borneo vs. Islands PCAs

###7. Mainland Borneo vs. Islands PCA (Manus, with offshore islands forced in)

```{r}
setwd("~/Desktop/thesis/complete data")
bornIx2 <- read.csv("manus_islandsB.csv")
bornIx2 <- na.omit(bornIx2)
bornIx2[,7:13] <- log(bornIx2[,7:13])

table(bornIx2$Locality)

pc11 <- princomp(bornIx2[,-(1:6)], cor=TRUE)
for (i in 1:7) {
  pc11$loadings[,i] <- (pc11$loadings[,i] * pc11$sdev[i])
}

print(summary(pc11),digits=2,loadings=pc11$loadings,cutoff=0)
round(pc11$sdev^2,2)

autoplot(pc11, data = bornIx2, colour = 'Locality', label = FALSE, label.size = 3) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (58.71%)") + ylab("PC2 (12.70%)")

autoplot(pc11, data = bornIx2, colour = 'Locality', label = FALSE, label.size = 3, frame = TRUE) + xlab("PC1 (58.71%)") + ylab("PC2 (12.70%)")
```

###8. Mainland Borneo vs. Islands PCA (Skull, with offshore islands forced in)

```{r}
bornIy <- y[which(!y$Locality == 'Mainland Sumatra'),]
bornIy <- bornIy[which(!bornIy$Subspecies == 'T. t. cervicalis'),]

table(bornIy$Locality)

pc6 <- princomp(bornIy[,-(1:8)], cor=TRUE)
for (i in 1:8) {
  pc6$loadings[,i] <- (pc6$loadings[,i] * pc6$sdev[i])
}
print(summary(pc6),digits=2,loadings=pc6$loadings,cutoff=0)
round(pc6$sdev^2,2)

autoplot(pc6, data = bornIy, colour = 'Locality', label = FALSE, label.size = 3) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (62.51%)") + ylab("PC2 (14.62%)")

autoplot(pc6, data = bornIy, colour = 'Locality', label = FALSE, label.size = 3, frame = TRUE) + xlab("PC1 (62.51%)") + ylab("PC2 (14.62%)")
```

##Mainland Sumatra vs. Islands PCAs

###9. Mainland Sumatra vs. Islands PCA (Manus, with offshore islands forced in)

```{r}
setwd("~/Desktop/thesis/complete data")
sumIx2 <- read.csv("manus_islandsS.csv")
sumIx2 <- na.omit(sumIx2)
sumIx2[,7:13] <- log(sumIx2[,7:13])

table(sumIx2$Subspecies)

pc9 <- princomp(sumIx2[,-(1:6)], cor=TRUE)
for (i in 1:7) {
  pc9$loadings[,i] <- (pc9$loadings[,i] * pc9$sdev[i])
}
print(summary(pc9),digits=2,loadings=pc9$loadings,cutoff=0)
round(pc9$sdev^2,2)

autoplot(pc9, data = sumIx2, colour = 'Locality', label = FALSE, label.size = 3) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (40.33%)") + ylab("PC2 (24.56%)")

autoplot(pc9, data = sumIx2, colour = 'Locality', label = FALSE, label.size = 3, frame = TRUE) + xlab("PC1 (40.33%)") + ylab("PC2 (24.56%)")
```

```{r}
#THIS IS WEIRD LEAVE THIS ONE OUT
setwd("~/Desktop/thesis/complete data")
sumIx22 <- read.csv("manus_islandsS2.csv")
sumIx2[,5:11] <- log(sumIx2[,5:11])

pc55 <- princomp(sumIx22[,-(1:4)], cor=TRUE)

for (i in 1:7) {
  pc55$loadings[,i] <- (pc55$loadings[,i] * pc55$sdev[i])
}

#print(summary(pc55),digits=2,loadings=pc55$loadings,cutoff=0)
#round(pc55$sdev^2,2)

#autoplot(pc55, data = sumIx22, colour = 'Locality', label = FALSE, label.size = 3) + geom_point(shape = 1, size = 2, colour = "black") 
#+ xlab("PC1 (40.33%)") + ylab("PC2 (24.56%)")

#autoplot(pc55, data = sumIx22, colour = 'Locality', label = FALSE, label.size = 3, frame = TRUE) 
#+ xlab("PC1 (40.33%)") + ylab("PC2 (24.56%)")
```


###10. Mainland Sumatra vs. Islands PCA (Skull, with offshore islands forced in)

```{r}
setwd("~/Desktop/thesis/complete data")
sumIy2 <- read.csv("cran_islandsS.csv")
sumIy2 <- na.omit(sumIy2)
sumIy2[,9:16] <- log(sumIy2[,9:16])

table(sumIy2$Subspecies)

pc10 <- princomp(sumIy2[,-(1:8)], cor=TRUE)

for (i in 1:8) {
  pc10$loadings[,i] <- (pc10$loadings[,i] * pc10$sdev[i])
}

print(summary(pc10),digits=2,loadings=pc10$loadings,cutoff=0)
round(pc10$sdev^2,2)

autoplot(pc10, data = sumIy2, colour = 'Locality', label = FALSE, label.size = 3) + geom_point(shape = 1, size = 2, colour = "black") + xlab("PC1 (48.87%)") + ylab("PC2 (17.02%)")

autoplot(pc10, data = sumIy2, colour = 'Locality', label = FALSE, label.size = 3, frame = TRUE) + xlab("PC1 (48.87%)") + ylab("PC2 (17.02%)")
```

##Mixed-Effect Models for PC1 and Latitude

###11. PC1 vs. Latitude (Manus)

```{r}
xpc1 <- pc1[["scores"]]
xpc1 <- xpc1[,1]
x$PC1 <- xpc1
ggplot(x, aes(y=PC1, x=lat)) + geom_point(aes(color = Locality)) + xlab("Latitude")
memx <- lmer(PC1 ~ lat + Locality + (1 | long), data = x, REML = FALSE)
summary(memx)
Anova(memx) 
```

###12. PC1 vs. Latitude (Skull)

```{r}
ypc1 <- pc2[["scores"]]
ypc1 <- ypc1[,1]
y$PC1 <- ypc1
ggplot(y, aes(y=PC1, x=lat)) + geom_point(aes(color = Locality)) + xlab("Latitude")
memy <- lmer(PC1 ~ lat + Locality + Sex + (1 | long), data = y, REML = FALSE)
summary(memy)
Anova(memy)
```

###13. PC1 vs. *Absolute value* of Latitude (Manus)

```{r}
x$abslat <- abs(x$lat)
ggplot(x, aes(y=PC1, x=abslat)) + geom_point(aes(color = Locality)) + xlab("Latitude")
memx2 <- lmer(PC1 ~ abslat + Locality + (1 | long), data = x, REML = FALSE)
summary(memx2)
Anova(memx2)
```

###14. PC1 vs. *Absolute value* of Latitude (Skull)

```{r}
y$abslat <- abs(y$lat)
ggplot(y, aes(y=PC1, x=abslat)) + geom_point(aes(color = Locality)) + xlab("Latitude")
memy2 <- lmer(PC1 ~ abslat + Locality + Sex + (1 | long), data = y, REML = FALSE)
summary(memy2)
Anova(memy2)
```

##Sexual Size Dimorphism

###DFA

```{r}
sex <- y[which(!y$Sex == 'sex unknown'),]
sex <- sex[which(!sex$Sex == ''),]
sex <- sex[which(!sex$Subspecies == 'kelabit'),]
sex$Sex <- droplevels(sex$Sex)
lin3 <- lda(sex[,-c(1:8, 17, 18)],grouping=sex$Sex)
ctraw3 <- table(sex$Sex, predict(lin3)$class)
ctraw3 #rows are actual count and columns are predicted count; ignore blank I column/row

# total percent correct
round(sum(diag(prop.table(ctraw3))),4)

table(sex$Sex)
```

```{r}
sexS <- sex[which(sex$Locality == 'Mainland Sumatra'),]
sexB <- sex[which(sex$Locality == 'Mainland Borneo'),]

table(sexS$Sex)
table(sexB$Sex)

linS <- lda(sexS[,-(1:8)],grouping=sexS$Sex)
ctrawS <- table(sexS$Sex, predict(linS)$class)
ctrawS #rows are actual count and columns are predicted count; ignore blank I column/row

# total percent correct
round(sum(diag(prop.table(ctrawS))),4)

linB <- lda(sexB[,-(1:8)],grouping=sexB$Sex)
ctrawB <- table(sexB$Sex, predict(linB)$class)
ctrawB #rows are actual count and columns are predicted count; ignore blank I column/row

# total percent correct
round(sum(diag(prop.table(ctrawB))),4)
```

###t-test on PC1 by island

```{r}
pcs <- princomp(sexS[,-(1:8)], cor=TRUE)
for (i in 1:8) {
  pcs$loadings[,i] <- (pcs$loadings[,i] * pcs$sdev[i])
}
print(summary(pcs),digits=4,loadings=pcs$loadings,cutoff=0)
sexS$PC1 <- pcs[["scores"]][,1]

autoplot(pcs, data = sexS, colour = 'Sex', label = FALSE, label.size = 3) + ggtitle("sumatra")

pcb <- princomp(sexB[,-(1:8)], cor=TRUE)
for (i in 1:8) {
  pcb$loadings[,i] <- (pcb$loadings[,i] * pcb$sdev[i])
}
print(summary(pcb),digits=4,loadings=pcb$loadings,cutoff=0)
sexB$PC1 <- pcb[["scores"]][,1]

autoplot(pcb, data = sexB, colour = 'Sex', label = FALSE, label.size = 3) + ggtitle("borneo")

t.test(PC1 ~ Sex, sexS)

aovS <- aov(sexS$PC1 ~ sexS$Sex)
summary(aovS)

t.test(PC1 ~ Sex, sexB)

aovB <- aov(sexB$PC1 ~ sexB$Sex)
summary(aovB)

pcov <- princomp(sex[,-(1:8)], cor=TRUE)
autoplot(pcov, data = sex, colour = 'Sex', label = FALSE, label.size = 3) + ggtitle("overall")
```


```{r}
setwd("~/Desktop/thesis/complete data")
cpl <- read.csv("cran.csv")
cpl <- na.omit(cpl[,c(1:4,8,9)])
cpl <- cpl[which(!cpl$Sex == 'sex unknown'),]
t.test(CPL ~ Sex, cpl)

cplB <- cpl[which(!cpl$Locality == 'Mainland Borneo'),]
cplS <- cpl[which(!cpl$Locality == 'Mainland Sumatra'),]

t.test(CPL ~ Sex, cplB)
t.test(CPL ~ Sex, cplS)
```

###t-test on PC1

```{r}
t.test(PC1 ~ Sex, sex)
```

###MANOVA on all variables

```{r}
#check multivariate normality: get online function
source("http://www.reuningscherer.net/STAT660/R/CSQPlot.r.txt")

for (i in 1:nlevels(sex$Sex)){
  CSQPlot(sex[sex$Sex == levels(sex$Sex)[i], c(9:16)],label= levels(sex$Sex)[i])
}

#satisfies this assumption for MANOVA!
```

```{r}
responsevar <- as.numeric(sex[,8])
for (i in 9:16) responsevar <- cbind(responsevar,as.numeric(sex[,i]))
```

```{r}
surveyaov <- manova(responsevar ~ sex$Sex)
summary.aov(surveyaov)
#summary.manova(surveyaov, tol=0)   #Default is Pillai's trace
#summary.manova(surveyaov,test="Wilks", tol=0)
#summary.manova(surveyaov,test="Roy", tol=0)
```
