---
title: "b&w figs"
author: "Maya Juman"
date: "May 8, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#load relevant packages
library(ggplot2)
library(ggfortify)
library(plyr)
library(gridExtra)
```

```{r}
#set up data
setwd("~/Desktop/thesis/complete data")
x <- read.csv("manus.csv")
y <- read.csv("cran.csv")
x[,7:22] <- log(x[,7:22])
y[,9:27] <- log(y[,9:27])

x <- x[,colSums(is.na(x)) < 20]
x <- na.omit(x)
x <- x[,-15]

y <- y[,colSums(is.na(y)) < 20]
y <- na.omit(y)

m <- c("grey90", "grey60", "black")
n <- c("grey80", "black")
```

##1. MAP

```{r}
library("rnaturalearth")
library("rnaturalearthdata")
library("ggspatial")
world <- ne_countries(scale = "medium", returnclass = "sf")
setwd("~/Desktop/thesis/gis")
sites <- read.csv("R_locdata.csv")
sites <- sites[-c(6,87:89),]
tsites <- read.csv("GIS_types.csv")
tsites <- tsites[-12,]

map <- ggplot(data = world) +
    geom_sf() + 
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), panel.grid.major = element_line(colour = "transparent"), panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    geom_point(data = sites, aes(x = Longitude, y = Latitude, shape = subsp.), size = 3, fill = "black") + 
  coord_sf(xlim = c(96,119), ylim = c(-6.5,9), expand = FALSE) + xlab("Longitude") + ylab("Latitude") + 
  scale_shape_manual(values = c(21,24), 
                     guide = guide_legend(label.theme = element_text(angle = 0, 
                                                                     face = "italic", size=12))) + 
  theme(legend.title = element_blank(), 
        legend.box.background = element_rect(color="black", size=0.4)) + 
  theme(legend.position = c(0.89,0.08),
        legend.spacing.x = unit(1, 'mm'), legend.key.size = unit(5, 'mm'), legend.background = element_rect(color = "white", linetype = "solid", size=1)) +
  annotation_scale(location = "bl", width_hint = 0.18) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.5, "in"), pad_y = unit(0.3, "in"), 
        height = unit(1, "cm"),
        width = unit(1, "cm"),
        style = north_arrow_fancy_orienteering)

col <- c("blue", "cyan", "darkgoldenrod", "coral", "chartreuse", "lightseagreen", "gray", "snow", "forestgreen", "deeppink", "darkorchid1", "plum1", "yellow", "navajowhite", "firebrick1")

theme_set(theme_bw())
rmap <- ggplot(data = world) +
    geom_sf() + 
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), panel.grid.major = element_line(colour = "transparent"), panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    geom_point(data = sites, aes(x = Longitude, y = Latitude, fill = subsp.), colour = "black", size = 2.5, shape = 21) + coord_sf(xlim = c(96,119), ylim = c(-6.5,9), expand = FALSE) + xlab("Longitude") + ylab("Latitude") + scale_fill_manual(values = col, guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=12))) + theme(legend.title = element_blank(), 
        legend.box.background = element_rect(color="black", size=0.4)) + 
  theme(legend.position = "right",
        legend.spacing.x = unit(1, 'mm'), legend.key.size = unit(5, 'mm'), legend.background = element_rect(color = "white", linetype = "solid", size=1)) + annotation_scale(location = "br", width_hint = 0.15) + geom_point(data = tsites, aes(x = Longitude, y = Latitude, shape = looked.at.it.), colour = "black", size = 1) + scale_shape_manual(values = c(17,20))


#use ppt or illustrator to format arrow, legend, 4 type labels, and 4 type stars
ggsave(rmap,file="Fig.1rev2.tiff",width=200, height=120, units="mm", dpi=600)
```

###2. All Subspecies PCA (Manus)

```{r}
pc1 <- princomp(x[,-(1:6)], cor=TRUE)
x1 <- x
x1$pc1 <- pc1$scores[,1]
x1$pc2 <- pc1$scores[,2]

#2A
a2 <- ggplot(data = x1, aes(x=pc1, y=pc2, group=Subspecies)) + geom_point(aes(shape=Subspecies), size=3) + scale_shape_manual(values = c(0,1,2,15,18,7,8,9,10,16,12), guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=9))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (50.66%)") + ylab("PC2 (15.35%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank(), legend.box.background = element_rect(color="black", size=0.5)) + theme(text = element_text(size=13)) + xlim(-4.5,6.3) + ylim(-4.3,2.85) + theme(legend.position = c(0.825, 0.185), legend.spacing.x = unit(2.5, 'mm'), legend.key.size = unit(3, 'mm'))
a2

ggsave(a2, file="2a.jpg", width=120, height=120, units = "mm")

find_hull <- function(x1) x1[chull(x1$pc1, x1$pc2), ]
hulls <- ddply(x1, "Locality", find_hull)

#2B
b2 <- ggplot(data = x1, aes(x=pc1, y=pc2, group=Locality)) + 
  geom_point(aes(shape=Locality), size=3) + 
  scale_shape_manual(values = c(2,10,20)) + 
  geom_vline(aes(xintercept = 0), size=0.25) + 
  scale_linetype_manual(values=c(2,3,1)) + 
  geom_hline(aes(yintercept = 0), size=0.25) + 
  theme_bw() + 
  xlab("PC1 (50.66%)") + ylab("PC2 (15.35%)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  scale_fill_manual(values=m) + 
  geom_polygon(data=hulls, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, 
                                          fill=Locality, linetype=Locality), 
               colour="black") + 
  theme(legend.position = c(0.79, 0.11), legend.spacing.x = unit(2.5, 'mm'), 
        legend.key.size = unit(6, 'mm')) + 
  ylim(-4,3) + xlim(-4.5,6.9) + 
  theme(legend.title = element_blank()) + 
  theme(text = element_text(size=13))
b2

ggsave(b2,file="2b.jpg",width=120, height=120, units="mm")

#grid.arrange(a2,b2,nrow=2)
```

###2C and 2D. All Subspecies PCA (Skull)

```{r}
pc2 <- princomp(y[,-(1:8)], cor=TRUE)
y1 <- y
y1$pc1 <- pc2$scores[,1]
y1$pc2 <- pc2$scores[,2]

#2C
a3 <- ggplot(data = y1, aes(x=pc1, y=pc2, group=Subspecies)) + geom_point(aes(shape=Subspecies), size=3) + scale_shape_manual(values = c(0,1,2,15,17,5,18,7,8,9,10,12,11), guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=9))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (65.58%)") + ylab("PC2 (13.92%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank()) + theme(text = element_text(size=13)) + theme(legend.position = c(0.825, 0.215), legend.spacing.x = unit(2.5, 'mm'), legend.key.size = unit(3, 'mm'), legend.box.background = element_rect(color="black", size=0.5)) + xlim(-6,9.5)
a3

ggsave(a3,file="3a.jpg",width=120, height=120, units="mm")

find_hull2 <- function(y1) y1[chull(y1$pc1, y1$pc2), ]
hulls2 <- ddply(y1, "Locality", find_hull2)

#2D
b3 <- ggplot(data = y1, aes(x=pc1, y=pc2, group=Locality)) + geom_point(aes(shape=Locality), size=3) + scale_shape_manual(values = c(2,10,20)) + geom_vline(aes(xintercept = 0), size=0.25) + scale_linetype_manual(values=c(2,3,1)) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (65.58%)") + ylab("PC2 (13.92%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_polygon(data=hulls2, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, linetype=Locality, fill=Locality), colour="black") + scale_fill_manual(values=m) + theme(legend.position = c(0.80, 0.11), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(6, 'mm')) + xlim(-6, 6.5) + ylim(-3.5,2.8) + theme(legend.title = element_blank()) + theme(text = element_text(size=13))
b3

ggsave(b3,file="3b.jpg",width=120, height=120, units="mm")
```

###3A. Mainland Borneo PCA (Manus)

```{r}
bornX <- x[which(x$Locality == 'Mainland Borneo'),]
pc3 <- princomp(bornX[,-(1:6)], cor=TRUE)
bornX1 <- bornX
bornX1$pc1 <- pc3$scores[,1]
bornX1$pc2 <- pc3$scores[,2]

a4 <- ggplot(data = bornX1, aes(x=pc1, y=pc2, group=Subspecies)) + geom_point(aes(shape=Subspecies), size=3) + scale_shape_manual(values = c(0,1,18,7,8,10,12), guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=9))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (54.63%)") + ylab("PC2 (12.56%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank()) + theme(text = element_text(size=13)) + xlim(-4.6,6.4) + theme(legend.position = c(0.868, 0.8475), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(4, 'mm'), legend.box.background = element_rect(color="black", size=0.5))
a4

ggsave(a4,file="4a.jpg",width=120, height=120, units="mm")
```

###3B. Mainland Borneo PCA (Skull)

```{r}
bornY <- y[which(y$Locality == 'Mainland Borneo'),]
pc4 <- princomp(bornY[,-(1:8)], cor=TRUE)
bornY1 <- bornY
bornY1$pc1 <- pc4$scores[,1]
bornY1$pc2 <- pc4$scores[,2]

b4 <- ggplot(data = bornY1, aes(x=pc1, y=pc2, group=Subspecies)) + geom_point(aes(shape=Subspecies), size=3) + scale_shape_manual(values = c(1,17,5,18,7,8,10,12,11), guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=9))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (58.52%)") + ylab("PC2 (17.50%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank()) + theme(text = element_text(size=13)) + theme(legend.position = c(0.8665, 0.169), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(3.5, 'mm'), legend.box.background = element_rect(color="black", size=0.5)) + xlim(-5.2,6.3)
b4

ggsave(b4,file="4b.jpg",width=120, height=120, units="mm")
```

###4A. Mainland Borneo vs. Islands PCA (Manus, with offshore islands forced in)

```{r}
setwd("~/Desktop/thesis/complete data")
bornIx2 <- read.csv("manus_islandsB.csv")
bornIx2 <- na.omit(bornIx2)
bornIx2[,7:13] <- log(bornIx2[,7:13])
pc11 <- princomp(bornIx2[,-(1:6)], cor=TRUE)
bx1 <- bornIx2
bx1$pc1 <- pc11$scores[,1]
bx1$pc2 <- pc11$scores[,2]

find_hull3 <- function(bx1) bx1[chull(bx1$pc1, bx1$pc2), ]
hulls3 <- ddply(bx1, "Locality", find_hull3)

a5 <- ggplot(data = bx1, aes(x=pc1, y=pc2, group=Locality)) + geom_point(aes(shape=Locality), size=3) + scale_shape_manual(values = c(2,20)) + geom_vline(aes(xintercept = 0), size=0.25) + scale_linetype_manual(values=c(2,1)) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (58.71%)") + ylab("PC2 (12.70%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_polygon(data=hulls3, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, linetype=Locality, fill=Locality), colour="black") + scale_fill_manual(values=n) + theme(legend.position = c(0.81, 0.08))  + xlim(-3.7, 4.55) + ylim(-2.2, 2.3) + theme(legend.title = element_blank(), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(6, 'mm')) + theme(text = element_text(size=13))
a5

ggsave(a5,file="5a.jpg",width=120, height=120, units="mm")
```

###4B. Mainland Borneo vs. Islands PCA (Skull, with offshore islands forced in)

```{r}
bornIy <- y[which(!y$Locality == 'Mainland Sumatra'),]
bornIy <- bornIy[which(!bornIy$Subspecies == 'cervicalis'),]
pc6 <- princomp(bornIy[,-(1:8)], cor=TRUE)
by1 <- bornIy
by1$pc1 <- pc6$scores[,1]
by1$pc2 <- pc6$scores[,2]

find_hull4 <- function(by1) by1[chull(by1$pc1, by1$pc2), ]
hulls4 <- ddply(by1, "Locality", find_hull4)

b5 <- ggplot(data = by1, aes(x=pc1, y=pc2, group=Locality)) + geom_point(aes(shape=Locality), size=3) + scale_shape_manual(values = c(2,20)) + geom_vline(aes(xintercept = 0), size=0.25) + scale_linetype_manual(values=c(2,1)) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (62.51%)") + ylab("PC2 (14.62%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_polygon(data=hulls4, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, linetype=Locality, fill=Locality), colour="black") + scale_fill_manual(values=n) + theme(legend.position = c(0.81, 0.08)) + xlim(-5.5, 6.2) + ylim(-3.2, 2.4) + theme(legend.title = element_blank(), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(6, 'mm')) + theme(text = element_text(size=13))
b5

ggsave(b5,file="5b.jpg",width=120, height=120, units="mm")
```

###4C. Mainland Sumatra vs. Islands PCA (Manus, with offshore islands forced in)

```{r}
setwd("~/Desktop/thesis/complete data")
sumIx2 <- read.csv("manus_islandsS.csv")
sumIx2 <- na.omit(sumIx2)
sumIx2[,7:13] <- log(sumIx2[,7:13])
pc9 <- princomp(sumIx2[,-(1:6)], cor=TRUE)
sx1 <- sumIx2
sx1$pc1 <- pc9$scores[,1]
sx1$pc2 <- pc9$scores[,2]

find_hull5 <- function(sx1) sx1[chull(sx1$pc1, sx1$pc2), ]
hulls5 <- ddply(sx1, "Locality", find_hull5)

#a6 <- ggplot(data = sx1, aes(x=pc1, y=pc2, group=Locality)) + geom_point(aes(shape=Locality), size=3) + scale_shape_manual(values = c(2,20)) + geom_vline(aes(xintercept = 0), size=0.25) + scale_linetype_manual(values=c(2,1)) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (40.33%)") + ylab("PC2 (24.56%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_fill_manual(values=n) + geom_polygon(data=hulls5, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, linetype=Locality, fill=Locality), colour="black") + theme(legend.position = c(0.21, 0.09)) + theme(legend.title = element_blank(), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(6, 'mm')) + theme(text = element_text(size=13)) + xlim(-4, 4)

a6 <- ggplot(data = sx1, aes(x=pc1, y=pc2, group=Locality)) + geom_point(aes(shape=Subspecies), fill="grey", size=4) + scale_shape_manual(values = c(18, 8, 5, 10)) + geom_vline(aes(xintercept = 0), size=0.25) + scale_linetype_manual(values=c(2,1)) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (40.33%)") + ylab("PC2 (24.56%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_fill_manual(values=n) + geom_polygon(data=hulls5, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, linetype=Locality, fill=Locality), colour="black") + theme(legend.position = "none") + theme(legend.title = element_blank(), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(6, 'mm')) + theme(text = element_text(size=13)) + xlim(-4, 4)

a6

setwd("~/Desktop/thesis/figs")
ggsave(a6,file="6areal.jpg",width=120, height=120, units="mm")
```

###4D. Mainland Sumatra vs. Islands PCA (Skull, with offshore islands forced in)

```{r}
setwd("~/Desktop/thesis/complete data")
sumIy2 <- read.csv("cran_islandsS.csv")
sumIy2 <- na.omit(sumIy2)
sumIy2[,9:16] <- log(sumIy2[,9:16])
pc10 <- princomp(sumIy2[,-(1:8)], cor=TRUE)
sy1 <- sumIy2
sy1$pc1 <- pc10$scores[,1]
sy1$pc2 <- pc10$scores[,2]

find_hull6 <- function(sy1) sy1[chull(sy1$pc1, sy1$pc2),]
hulls6 <- ddply(sy1, "Locality", find_hull6)

b6 <- ggplot(data = sy1, aes(x=pc1, y=pc2, group=Locality)) + geom_point(aes(shape=Locality), size=3) + scale_shape_manual(values = c(2,20)) + geom_vline(aes(xintercept = 0), size=0.25) + scale_linetype_manual(values=c(2,1)) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (48.87%)") + ylab("PC2 (17.02%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_fill_manual(values=n) + geom_polygon(data=hulls6, alpha=0.3, aes(x=pc1, y=pc2, group=Locality, linetype=Locality, fill=Locality), colour="black") + theme(legend.position = c(0.21, 0.09)) + xlim(-5, 5) + theme(legend.title = element_blank(), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(6, 'mm')) + theme(text = element_text(size=13))
b6

ggsave(b6,file="6br.jpg",width=120, height=120, units="mm")
```

